Think deep for reef resilience

### 2025-08-10 paper/paper_draft.qmd
Adjust text to reflect the new ECS values and quantification
Reflect quantification of low-high cluster timeseries differences using CID metric.

## Future DHW trajectories

Each Global Climate Model (GCM) shows similar relationships between reef clusters and
contextual analysis factors, thus further comparisons of these clusters for each analysis
factor are only shown for ACCESS-ESM1-5 (an intermediate Equilibrium Climate Sensitivity
scenario). Results for the additional GCMs are displayed in the supplementary material.
However, there are substantial differences in the longevity of reef resilience and
maintenance of coral cover post 2030 between the GCMs, thus coral cover timeseries are
displayed for each GCM (displayed clustered at management-area scale to preserve coarse
spatial differences in coral trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
Recovery phases are predominantly seen in the ‘high’ cover cluster. The ‘medium’ cluster,
and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing
proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern management area
does not see coral cover maintenance as high as the other management regions \[**TI:**
**Quantify this! How did you determine this and how large/small are the difference in coral
cover maintenance?**\]. Under the second lowest ECS projection (GFDL-CM4, 2.9°C) all
clusters in each management area exhibit largely declining future trajectories with periods
of recovery from 2030 to 2040. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which
results in intermediate levels of coral cover maintenance (between those of lower and higher
ECS GCMs). In Townsville-Whitsunday and Far Northern management areas, reefs in the high
cluster are able to maintain coral cover levels around 20% to 2050 under ACCESS-ESM1-5.
Under ACCESS-CM2 (4.72°C) and EC-Earth3-Veg (4.31°C) GCMs cover trajectories are largely
declining in all management areas.

Under each GCM, in all management areas, low, medium and high cluster reefs have low
differences between DHW trajectories (@fig-gcm-dhw-trajectories) \[**TI:** **I don't
understand this sentence - what is meant by reefs have low differences?**\]. The highest
peak DHW levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in
overall DHW timeseries occur under ACCESS-CM2.

@@

Comparing across GCMs, it is common for reefs to switch cluster level, including from low to
high and high to low cluster transitions (@sfig-gcm-bioregion-cluster-assignment). This is
expected as spatial differences in GCM DHW timeseries result in reefs being impacted in
different magnitudes relative to other reefs in their bioregion.

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median timeseries in each management area under each GCM. CID values are calculated for percentage cover timeseries (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) timeseries (@fig-gcm-dhw-trajectories). Complexity invariant distance represents the euclidean distance between timeseries, corrected by the complexity of the two timeseries. CID values for percentage cover and DHW timeseries are not directly comparable as the timeseries use different units.

@@

\[**BG: Need to add quantification results to text for timeseries differences.**\]

@@

Under all GCMs most reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover
maintenance into the future. Differences between low, medium and high reef clusters in cover
timeseries are most notable under these two GCMs. Maintainance of coral cover into the
future is not clearly linked to the Equilibrium Climate Sensitivity of each GCM used as
ACCESS-ESM1-5 has a higher ECS value than GFDL-CM4, with reefs experiencing large continuous
declines under GFDL-CM4. These large declines are likely due to the high peak thermal stress
levels above 10°C-weeks experienced in the first ten years of the timeseries under GFDL-CM4.

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider this uncertainty in future conditions when assessing possible biological
responses to change. This is supported by recent literature suggesting an effective approach
to capturing climate change uncertainty is to model responses using a range of GCMs
separately [@shepherd2018]. Despite large differences in coral cover across GCMs, the
patterns identified between low, medium and high clusters of reefs suggest that the
importance of depth refugia is consistent across scenarios.

>>>>

## Reef trajectories across futures

Each Global Climate Model (GCM) shows similar relationships between reef clusters and
contextual analysis factors, thus further comparisons of these clusters for each analysis
factor are only shown for ACCESS-ESM1-5 (an intermediate Equilibrium Climate Sensitivity
scenario [@sfig-ecs]). Results for the additional GCMs can be found at \[**BG: Insert
repository citation.**\]. However, there are substantial differences in the longevity of
reef resilience and maintenance of coral cover post 2030 between the GCMs, thus coral cover
timeseries are displayed for each GCM (displayed clustered at management-area scale to
preserve coarse spatial differences in cover trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
Recovery phases are predominantly seen in the ‘high’ cover cluster. The ‘medium’ cluster,
and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing
proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern management area has
lower CID distances between low and high cover clusters than in other management areas
[@tbl-cluster-cid], this is reflected in the low degree of cover maintainence
[@fig-gcm-cover-trajectories]. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which
results in intermediate levels of coral cover maintenance (between those of lower and higher
ECS GCMs) [@fig-gcm-cover-trajectories], reflected in intermediate CID distance values
within most management areas [@tbl-cluster-cid]. In Townsville-Whitsunday and Far Northern
management areas, reefs in the high cluster are able to maintain coral cover levels around
20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.66K), EC-Earth3-Veg (4.30K) and
GFDL-CM4 (4.1K) GCMs cover trajectories are largely declining in all management areas,
reflected in lower CID distance values than other GCMs [@tbl-cluster-cid].

Under each GCM, in all management areas, low, medium and high cluster reefs have low levels
of distinction between DHW trajectories (@fig-gcm-dhw-trajectories). The highest peak DHW
levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs
timeseries occur under ACCESS-CM2 [@fig-methods-dhws].

@@

Comparing across GCMs, it is common for reefs to switch cluster level, including from low to
high and high to low cluster transitions (@sfig-gcm-bioregion-cluster-assignment). This is
expected as spatial differences in GCM DHW timeseries result in reefs being impacted at
different relative magnitudes within their management area across GCMs.

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median timeseries in each management area under each GCM. CID values are calculated for percentage cover timeseries (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) timeseries (@fig-gcm-dhw-trajectories). Complexity invariant distance represents the euclidean distance between timeseries, corrected by the complexity of the two timeseries. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW timeseries are not directly comparable as the timeseries use different units.

@@

@@

Under all GCMs most reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover
maintenance into the future. Differences between low, medium and high reef clusters in cover
timeseries are most notable under these two GCMs [@tbl-cluster-cid;
@fig-gcm-cover-trajectories]. Maintainance of coral cover into the future is linked to the
Equilibrium Climate Sensitivity of the GCM used as levels of future recovery and timespan of
cover maintainance increase as GCM ECS increases [@fig-gcm-cover-trajectories]. Under
high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2) large declines and lower rates of
recovery in high cover clusters are common compared to lower-ECS GCMs.

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider uncertainty in future conditions when assessing possible biological responses to
change. This is supported by recent literature suggesting an effective approach to capturing
climate change uncertainty is to model responses using GCMs separately [@shepherd2018].
Despite large differences in coral cover across GCMs, the patterns identified between low,
medium and high clusters of reefs suggest that the relative importance of environmental and
biological factors such as depth refugia, connectivity and surface DHW levels is consistent
across scenarios.
